Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-30 paper.qmd
Changed augment arguments with new quollr version

## Compute error
error_df_two_curvy_abs <- augment(
  df_bin_centroids = df_bin_centroids_two_curvy,
  df_bin = df_bin_two_curvy,
  training_data = training_data_two_curvy,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "x")

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.1
)

@@

error_df_gau_abs <- augment(
  df_bin_centroids = df_bin_centroids_gau1,
  df_bin = df_bin_gau1,
  training_data = training_data_gau,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "x")

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = umap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x"
)

@@

error_df_gau_abs <- augment(
  df_bin_centroids = df_bin_centroids_gau1,
  df_bin = df_bin_gau1,
  training_data = training_data_gau,
  newdata = NULL,
  type_NLDR = "UMAP",
  col_start = "x")

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.1
)

@@

error_df_gau_abs <- augment(
  df_bin_centroids = df_bin_centroids_gau1,
  df_bin = df_bin_gau1,
  training_data = training_data_gau,
  newdata = NULL,
  type_NLDR = "PaCMAP",
  col_start = "x")

@@

```{r}
#| label: error-model-one-dens-clust
## Compute error
error_df_one_curvy_abs <- augment(
  df_bin_centroids = df_bin_centroids_dens_clust,
  df_bin = df_bin_dens_clust,
  training_data = one_c_shaped_data,
  newdata = NULL,
  type_NLDR = "tSNE",
  col_start = "x")

@@

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("pc", 1:50)

@@

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  num_x_bins = num_bins_x_pbmc,
  col_start = "pc")

@@

<!-- PaCMAP layout with author's suggested parameter choice-->
```{r}
#| label: read-mnist
## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

@@

algo_obj_mnist <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_mnist, 
  nldr_data = tsne_mnist2, 
  num_x_bins = num_bins_x_mnist,
  col_start = "PC")

@@

## Compute error
error_df <- augment(
  df_bin_centroids = df_bin_centroids_mnist, 
  df_bin = df_bin_mnist, 
  training_data = training_data_mnist, 
  newdata = NULL, 
  type_NLDR = "tSNE", 
  col_start = "PC")

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "The $10\\text{-}D$ model error in \\gD{} layout of the MNIST digit 1 data shows a pattern. Most low model errors are distributed along the big non-linear cluster, while most large model errors are distributed along the small cluster. The images associated with large model errors shows different patterns of digit 1, some inside (e) the non-linear structure and others outside (d). Along the non-linear cluster, the angle of digit 1 changes (a-c)."
#| fig-pos: H
#| fig-width: 7
#| fig-height: 5
#| out-width: 90%

>>>>

## Compute error
error_df_two_curvy_abs <- augment(
  highd_data = training_data_two_curvy,
  model_2d = df_bin_centroids_two_curvy,
  model_highd = df_bin_two_curvy)

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  q = 0.1
)

@@

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = umap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE
)

@@

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  q = 0.1
)

@@

error_df_gau_abs <- augment(
  highd_data = training_data_gau,
  model_2d = df_bin_centroids_gau1,
  model_highd = df_bin_gau1)

@@

```{r}
#| label: error-model-one-dens-clust
## Compute error
error_df_one_curvy_abs <- augment(
  highd_data = one_c_shaped_data,
  model_2d = df_bin_centroids_dens_clust,
  model_highd = df_bin_dens_clust)

@@

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| label: published-pbmc
## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
names(training_data_pbmc) <- paste0("x", 1:50)

@@

algo_obj_pbmc <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_pbmc, 
  nldr_data = umap_pbmc, 
  num_x_bins = num_bins_x_pbmc,
  col_start = "x")

@@

<!-- PaCMAP layout with author's suggested parameter choice-->
```{r}
#| label: read-mnist
## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
names(training_data_mnist) <- paste0("x", 1:NCOL(training_data_mnist))

training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

@@

algo_obj_mnist <- gen_nldr_vis_algo_obj(
  high_d_data = training_data_mnist, 
  nldr_data = tsne_mnist2, 
  num_x_bins = num_bins_x_mnist,
  col_start = "x")

@@

## Compute error
error_df <- augment(
  highd_data = training_data_mnist,
  model_2d = df_bin_centroids_mnist,
  model_highd = df_bin_mnist)

@@

```{r}
#| label: fig-model-error-mnist
#| fig-cap: "The $10\\text{-}D$ model error in \\gD{} layout of the MNIST digit 1 data shows a pattern. Most low model errors are distributed along the big non-linear cluster, while most large model errors are distributed along the small cluster. The images associated with large model errors shows different patterns of digit 1, some inside (e) the non-linear structure and others outside (d). Along the non-linear cluster, the angle of digit 1 changes (a-c)."
#| fig-pos: H
#| fig-width: 10
#| fig-height: 7
#| out-width: 90%
